
---
title: "Crime Data Analysis"
author: "Prajwal KC"
output: html_document
---

## Introduction

Crime data analysis is essential for identifying patterns, trends, and areas requiring law enforcement attention. This report explores a publicly available U.S. crime dataset sourced from Kaggle. The goal is to analyze crime trends over time, examine state-wise distributions, and conduct an in-depth analysis of Los Angeles, California.

By understanding how crimes vary by time, region, and demographics, data analysts can provide valuable insights for crime prevention, policy planning, and public safety improvements.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Dataset Description

The dataset includes nationwide crime reports across several years and states in the U.S. Key features include:
- `Year`: The year the crime was reported
- `State` and `City`: The location of the crime
- `Crime Type`, `Weapon`, `Crime Solved`: Information about the crime event
- `Victim Age`, `Victim Sex`, `Victim Race`, `Relationship`: Victim demographics

The dataset contains thousands of observations, offering sufficient depth for statistical and trend analysis.

## Load the Dataset

```{r load-data}
# Load the dataset (update the path if needed)
# Example: crimedata <- read.csv("path/to/US_Crime_DataSet.csv")
crimedata <- US_Crime_DataSet_csv
```

## Crime Types Overview

```{r crime-types}
unique(crimedata$`Crime Type`)
table(crimedata$`Crime Type`)
```

## Crime Trends Over Time

```{r crime-trends}
total_crime <- crimedata %>%
  group_by(Year) %>%
  summarize(count = n())

ggplot(total_crime, aes(x = Year, y = count)) +
  geom_line(color = "steelblue", size = 1) +
  scale_x_continuous(breaks = seq(min(total_crime$Year), max(total_crime$Year), by = 5)) +
  labs(
    x = "Year",
    y = "Total Number of Crimes",
    title = "Crime Trends Over Time"

The following plot shows how the total number of crimes reported has changed over the years.
  ) +
  theme_minimal()
```

## Statewise Crime Distribution

```{r statewise-crime}
statewise <- crimedata %>%
  group_by(State) %>%
  summarize(number = n())

ggplot(statewise, aes(x = reorder(State, -number), y = number)) +
  geom_col(fill = "coral") +
  coord_flip() +
  labs(title = "Total Crimes by State"

This bar chart reveals which states have the highest and lowest number of reported crimes., x = "State", y = "Number of Crimes") +
  theme_minimal()
```

## Los Angeles Crime Focus

```{r la-analysis}
LA <- crimedata %>% filter(City == "Los Angeles")

LA %>% group_by(`Crime Solved`) %>% summarize(count = n())

# Insight: This shows how many crimes were solved versus unsolved in LA.
LA %>% group_by(`Victim Sex`) %>% summarize(count = n())
LA %>% group_by(`Victim Race`) %>% summarize(count = n())
```

## Victim Age Group Distribution

```{r age-group}
agegroup <- LA %>%
  mutate(agegroup = case_when(
    `Victim Age` < 5 ~ "Under 5",
    `Victim Age` >= 5 & `Victim Age` < 18 ~ "5-17",
    `Victim Age` >= 18 & `Victim Age` < 50 ~ "18-49",
    `Victim Age` >= 50 ~ "50+"
  ))

agegroup %>%
  group_by(agegroup) %>%
  summarize(count = n())
```

## Common Weapons Used

```{r weapons-used}
weapon <- c("Handgun", "Knife", "Blunt Object", "Firearm", "Rifle", "Shotgun")

ggplot(data = subset(LA, Weapon %in% weapon), aes(x = Weapon)) +
  geom_bar(fill = "steelblue") +
  scale_y_continuous(breaks = seq(0, 25000, by = 5000)) +
  labs(title = "Weapons Used in Crimes"

This bar chart highlights the most common weapons used in Los Angeles crime reports., x = "Weapon", y = "Count") +
  theme_minimal()
```

## Linear Regression on Total Crimes

```{r regression-total-crime}
linear <- lm(count ~ Year, data = total_crime)
summary(linear)
total_crime$linearpred <- predict(linear, newdata = total_crime)

ggplot(total_crime, aes(x = Year, y = count)) +
  geom_line(color = "blue") +
  geom_line(aes(y = linearpred), color = "red", linetype = "dashed") +
  labs(title = "Linear Regression on Total Crimes"

This regression model helps evaluate whether crime numbers have increased or decreased over the years., x = "Year", y = "Count")
```

## Conclusion

This analysis uncovered several valuable insights from the U.S. crime dataset:

- Crimes have shown varying trends over the years, with some periods of increase or decline.
- Certain states contribute disproportionately to national crime totals.
- Los Angeles, as a major urban area, shows distinct crime patterns — including common weapons used and age demographics.
- Linear regression analysis shows time-based trends which could help in forecasting and prevention strategies.

Overall, this type of analysis can support policymakers, safety organizations, and researchers in making data-informed decisions. into how crime patterns vary by time, location, and demographic factors. The linear regression model suggests trends that can be further explored for predictive insights. The city-specific breakdown helps focus policy and safety measures.
